Solving Collective Spin Hamiltonians with Decoherence

### 2018-08-27 soc_squeezing/cumulant_expansions.tex
replace explanations for dynamics with TVF, TAT by that for the general case

Previous work has shown how to realize various spin-squeezing
Hamiltonians with ultracold atomic systems.  While it is
straightforward to simulate the dynamics generated by these
Hamiltonians in the context of a closed quantum system, here we try to
develop an understanding of their the squeezing behavior in the
presence of decoherence.  In particular, we will be considering
dynamics induced by the collective one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_z^2,
  &&
  H_{\t{TAT}}^{\p{z,y}} = \chi \p{S_z^2 - S_y^2},
  &&
  H_{\t{TVF}} = \chi S_z^2 + \Omega S_x,
\end{align}
for spins which decay in an uncorrelated manner at a rate $\gamma$.
The evolution of any Heisenburg operator $\O$ with a Hamiltonian $H$
is given by
\begin{align}
  \f{d}{dt} \O
  = i\sp{H,\O}_- + \gamma\D_-\p{\O},
  &&
  \D_-\p{\O} \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+},
  \label{eq:EOM}
\end{align}
where $\sp{A,B}_\pm\equiv AB\pm BA$ and $j$ indexes an individual
spin.  Our measure of squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by components
$\bk{S_\alpha}$ of the mean spin vector in addition to collective
spin-spin correlators of the form $\bk{S_\alpha S_\beta}$; our task is
therefore to compute these expectation values in all scenarios of
interest.

@@

The equations of motion for $S_\mu$, $S_\mu S_z$, and $S_\mu S_\nu$
(i.e.~where $\mu,\nu\in\set{+1,-1}$) under the OAT Hamiltonian
$H_{\t{OAT}}=\chi S_z^2$ are determined by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\chi \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n},
\end{align}
where we can use identity \eqref{eq:comm_z_mu} to simplify
\begin{align}
  \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  = \sp{S_z^2, S_+^\ell S_-^m}_- S_z^n
  = S_+^\ell S_-^m \p{\sp{\ell-m}^2 + 2\sp{\ell-m} S_z} S_z^n,
  \label{eq:S_z^2_comm}
\end{align}
and the effect of decoherence through $\D_-\p{S_+^\ell S_-^m S_z^n}$ is
worked out in Appendix \ref{sec:decoherence}.  In order to compute
squeezing of the OAT Hamiltonian in the presence of single-spin decay,
we therefore need to solve the systems of equations defined by
\begin{align}
  \f{d}{dt} \p{S_\mu S_z^n}
  = i\chi S_\mu \p{1 + 2\mu S_z} S_z^n
  + \gamma S_\mu \sp{\p{S - \delta_{\mu,-1} + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - \f12 S_z^n}
  \label{eq:evolution_OAT_single}
\end{align}
and
\begin{multline}
  \f{d}{dt} \p{S_\mu S_\nu S_z^n}
  = i\chi S_\mu S_\nu \p{\sp{\mu+\nu}^2 + 2 \sp{\mu+\nu} S_z} S_z^n \\
  + \gamma S_\mu S_\nu
  \sp{\p{S - \delta_{\mu,-1} - \delta_{\nu,-1} + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - S_z^n},
  \label{eq:evolution_OAT_pair}
\end{multline}
for $\mu\ge\nu$.  The initial conditions for these systems of
equations can be determined from the expansions
\begin{align}
  S_z^n = \sum_{m=-S}^S m^n \op{m},
  &&
  S_\mu = \sum_{m=-S}^S g_\mu\p{m} \op{m+\mu}{m},
\end{align}
with
\begin{align}
  g_\mu\p{S,m} \equiv \sqrt{\p{S-\mu m}\p{S+\mu m+1}},
  \label{eq:transverse_elem}
\end{align}
and the initial state
\begin{align}
  \ket{X} = 2^{-N/2} \sum_{m=-S}^S { N \choose S+m }^{1/2} \ket{m},
\end{align}
from which we can determine
\begin{align}
  \bk{S_\mu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S}^S m^n { N \choose S+m } \abs{m - \mu S},
\end{align}
and
\begin{align}
  \bk{S_\mu S_\nu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S+\delta_{\mu,-1}+\delta_{\nu,-1}}
  ^{S-\delta_{\mu,1}-\delta_{\nu,1}}
  m^n \sp{{ N \choose S+m+\mu+\nu } { N \choose S+m }}^{1/2}
  g_\mu\p{S,m+\nu} g_\nu\p{S,m}.
\end{align}

\subsection{Operator imbalance, conservation laws, and time evolution}

For convenience of language, for collective spin operators of the form
$S_+^\ell S_-^m S_z^n$ we respectively define the {\it upper}, {\it
  lower}, and {\it net imbalance}
\begin{align}
  \I_+\p{S_+^\ell S_-^m S_z^n} \equiv \ell,
  &&
  \I_-\p{S_+^\ell S_-^m S_z^n} \equiv m,
  &&
  \I\p{S_+^\ell S_-^m S_z^n} \equiv \ell - m.
\end{align}
In the case of one-axis twisting, both upper and lower (and net)
imbalances are {\it conserved}, meaning that operators with upper and
lower imbalances $\p{\ell,m}$ depend only on other operators with the
same imbalances.  We can therefore divide the task of operator time
evolution into sectors of constant imbalances, which greatly reduces
computational overhead.  Specifically, operator time evolution now
involves evolving ``operator vectors''
\begin{align}
  \v V_{\t{OAT}}^{(\ell,m)}
  \equiv S_+^\ell S_-^m \sum_{n\ge0} S_z^n \uv x_n,
\end{align}
which in practice must be truncated at some maximal $n=M$.  In the
case of OAT, the time derivative of $S_+^\ell S_-^m S_z^n$ depends on
$S_+^\ell S_-^m S_z^{n+1}$ which implies that we can only trust the
value of $S_+^\ell S_-^m S_z^n$ up to the time at which the initial
value of $S_+^\ell S_-^m S_z^M$ has (conservatively) an $\O\p{1}$
contribution to $S_+^\ell S_-^m S_z^n$.

In order to compute squeezing from OAT, we need only need to evolve
$\v V_{\t{OAT}}^{(1,0)}$, $\v V_{\t{OAT}}^{(0,1)}$,
$\v V_{\t{OAT}}^{(1,1)}$, and $\v V_{\t{OAT}}^{(2,0)}$.  The values of
axial spin operators $S_z$ and $S_z^2$ are known exactly at all times
from \eqref{eq:Sz_OAT}.

\section{Transverse-field and two-axis twisting}

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_-^q S_z^r.
  \label{eq:H_general}
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_-^q S_z^r, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n},
\end{align}
where the commutator is given in Appendix \ref{sec:comm_general}, and
the decoherence operator in Appendix \ref{sec:decoherence}.  We will
be considering evolution of the initial states
$\ket{X}:S_x\ket{X}=S\ket{X}$ and $\ket{-Z}:S_z\ket{-Z}=-S\ket{-Z}$,
for which initial conditions are given by
\begin{multline}
  \bk{X|S_+^\ell S_-^m S_z^n|X}
  =  2^{-N} \sum_{k=-S+m}^{S-\ell} k^n
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}
\end{multline}
and
\begin{multline}
  \bk{-Z|S_+^\ell S_-^m S_z^n|-Z}
  = \p{-S}^n 2^{-N} \sum_{k=-S+m}^{S-\ell}
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}.
\end{multline}

\subsection{Transverse-field Hamiltonian}

The transverse-field (TVF) twisting Hamiltonian with unit twisting
strength is
\begin{align}
  H_{\t{TVF}}
  = S_z^2 + \Omega S_x
  = S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-.
\end{align}
Unlike in the case of OAT, we now cannot divide all operators of the
form $S_+^\ell S_-^m S_z^n$ into independent sectors of constant
imbalances $\ell,m$.  We therefore need to keep track of all operators
in one vector $\v V_{\t{TVF}}$, which in turn requires imposing an
ordering on all operators of the form $S_+^\ell S_-^m S_z^n$.  Such an
ordering is provided by the one-to-one map $u:\N_0^3\to\N_0$, where$
\N_0$ denotes the set of non-negative integers, given by
\begin{align}
  u\p{\ell,m,n} = B\p{T_{\ell mn}} + R\p{T_{\ell mn},\ell} + m,
  \label{eq:vec_map}
\end{align}
where $T_{\ell mn}\equiv\ell+m+n$ for shorthand and
\begin{align}
  B\p{T}
  \equiv \abs{\set{\p{\ell,m,n}\in\N_0^3 : T_{\ell mn}<T}}
  = \f16 T \p{T+1} \p{T+2},
\end{align}
\begin{align}
  R\p{T,L}
  \equiv \abs{\set{\p{\ell,m,n}\in\N_0^3 : T_{\ell mn}=T , \ell<L}}
  = \f12 \ell \p{2 T - \ell + 3}.
\end{align}
In words, $u\p{\ell,m,n}$ sorts the collection non-negative integers
$\p{\ell,m,n}$ first by their sum $T_{\ell mn}$, and then in turn by
the values of $\ell$ and $m$.  Fixing a maximal operator weight $M$,
the vector keeping track of all operators is then
\begin{align}
  \v V_{\t{TVF}}
  \equiv \sum_{\ell+m+n\le M} S_+^\ell S_-^m S_z^n \uv x_{u\p{\ell,m,n}}.
\end{align}

\subsection{Two-axis twisting in-plane with the decoherence axis}

The two-axis twisting Hamiltonian one axis oriented along the
decoherence axis is
\begin{align}
  H_{\t{TAT}}^{\p{z,y}}
  = S_z^2 - S_y^2
  = S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z.
\end{align}
While upper and lower imbalances are not conserved by this
Hamiltonians, the parity of the net imbalance, i.e.~$\ell-m~\t{mod}~2$
for an operator $S_+^\ell S_-^m S_z^n$, is conserved.  There are
therefore two sectors of operators which we can evolve independently
to save computational overhead.  To keep track of these operators in
one-dimensional vectors, we define the onto maps $v_p:\N_0^3\to\N_0$
for $p\in\set{0,1}$ given by
\begin{align}
  v_p\p{\ell,m,n} = B_p\p{T_{\ell mn}} + R_p\p{T_{\ell mn},n} + m,
\end{align}
where
\begin{align}
  B_p\p{T}
  &\equiv \abs{\set{\p{\ell,m,n}\in\N_0^3 :
      T_{\ell mn}<T, \ell-m=p~\t{mod}~2}} \\
  &= \f12 B\p{T} + \f14 \p{T+1} \p{T~\t{mod}~2} \p{1-2p},
\end{align}
\begin{align}
  R_p\p{T,N}
  &\equiv \abs{\set{\p{\ell,m,n}\in\N_0^3 :
      T_{\ell mn}=T, \ell-m=p~\t{mod}~2, n < N }} \\
  &= \floor{\f{n}{2}}
  \p{T - \floor{\f{n}{2}} + 2 - \p{T+p~\t{mod}~2}}.
\end{align}
The map $v_p\p{\ell,m,n}$ sorts the collection non-negative integers
$\p{\ell,m,n}$ with $\ell-m=p~\t{mod}~2$ first by their sum
$T_{\ell mn}$, and then in turn by the values of $n$ and $m$.  Fixing
a maximum operator weight $M$, the vectors keeping track of all
operators are then
\begin{align}
  \v V_{\t{TAT}}^{\p{z,y,p}}
  \equiv \sum_{\substack{\ell,m,n\le M\\\ell-m=p~\t{mod}~2}}
  S_+^\ell S_-^m S_z^n \uv x_{v_p\p{\ell,m,n}}.
\end{align}

\subsection{Two-axis twisting transverse to the decoherence axis}

The two-axis twisting Hamiltonian with both axes transverse to the
decoherence axis is
\begin{align}
  H_{\t{TAT}}^{y,x}
  = S_y^2 - S_x^2
  = - \f12 S_+^2 - \f12 S_-^2.
\end{align}
Just as with the previous TAT Hamiltonian, the parity of the net
imbalance of operators is conserved by $H_{\t{TAT}}^{y,x}$, implying
that there are two independent sectors of operators to evolve.

>>>>

Previous work has shown how to realize various spin-squeezing
Hamiltonians with ultracold atomic systems.  While it is
straightforward to simulate the dynamics generated by these
Hamiltonians in the context of a closed quantum system, here we try to
develop an understanding of their the squeezing behavior in the
presence of decoherence.  In particular, we will be considering
dynamics induced by the collective one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_z^2,
  &&
  H_{\t{TAT}}^{z,y} = \chi \p{S_z^2 - S_y^2},
  &&
  H_{\t{TVF}} = \chi S_z^2 + \Omega S_x,
\end{align}
for spins which decay in an uncorrelated manner at a rate $\gamma$.
The evolution of any Heisenburg operator $\O$ with a Hamiltonian $H$
is given by
\begin{align}
  \f{d}{dt} \O
  = i\sp{H,\O}_- + \gamma\D_-\p{\O},
  &&
  \D_-\p{\O} \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+},
  \label{eq:EOM}
\end{align}
where $\sp{A,B}_\pm\equiv AB\pm BA$ and $j$ indexes an individual
spin.  Our measure of squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by components
$\bk{S_\alpha}$ of the mean spin vector in addition to collective
spin-spin correlators of the form $\bk{S_\alpha S_\beta}$; our task is
therefore to compute these expectation values in all scenarios of
interest.

@@

The equations of motion for $S_\mu$, $S_\mu S_z$, and $S_\mu S_\nu$
(i.e.~where $\mu,\nu\in\set{+1,-1}$) under the OAT Hamiltonian
$H_{\t{OAT}}=\chi S_z^2$ are determined by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\chi \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n},
\end{align}
where we can use identity \eqref{eq:comm_z_mu} to simplify
\begin{align}
  \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  = \sp{S_z^2, S_+^\ell S_-^m}_- S_z^n
  = S_+^\ell S_-^m \p{\sp{\ell-m}^2 + 2\sp{\ell-m} S_z} S_z^n,
  \label{eq:S_z^2_comm}
\end{align}
and the effect of decoherence through $\D_-\p{S_+^\ell S_-^m S_z^n}$ is
worked out in Appendix \ref{sec:decoherence}.  In order to compute
squeezing of the OAT Hamiltonian in the presence of single-spin decay,
we therefore need to solve the systems of equations defined by
\begin{align}
  \f{d}{dt} \p{S_\mu S_z^n}
  = i\chi S_\mu \p{1 + 2\mu S_z} S_z^n
  + \gamma S_\mu \sp{\p{S - \delta_{\mu,-1} + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - \f12 S_z^n}
\end{align}
for $\mu\in\set{+1,-1}$, and
\begin{align}
  \f{d}{dt} \p{S_+^2 S_z^n}
  = 4i\chi S_\mu S_\nu \p{1 + S_z} S_z^n
  + \gamma S_\mu S_\nu \sp{\p{S + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - S_z^n},
\end{align}
while $d/dt\p{S_\pm S_\mp}=0$, so
\begin{align}
  \bk{S_\pm S_\mp} = \bk{S_\pm S_\mp}_{t=0} e^{-\gamma t}.
\end{align}
The initial conditions for these systems of equations can be
determined from the expansions
\begin{align}
  S_z^n = \sum_{m=-S}^S m^n \op{m},
  &&
  S_\mu = \sum_{m=-S}^S g_\mu\p{m} \op{m+\mu}{m},
\end{align}
with
\begin{align}
  g_\mu\p{S,m} \equiv \sqrt{\p{S-\mu m}\p{S+\mu m+1}},
  \label{eq:transverse_elem}
\end{align}
and the initial state
\begin{align}
  \ket{X} = 2^{-N/2} \sum_{m=-S}^S { N \choose S+m }^{1/2} \ket{m},
\end{align}
from which we can determine
\begin{align}
  \bk{S_\mu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S}^S m^n { N \choose S+m } \abs{m - \mu S},
\end{align}
and
\begin{align}
  \bk{S_\mu S_\nu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S+\delta_{\mu,-1}+\delta_{\nu,-1}}
  ^{S-\delta_{\mu,1}-\delta_{\nu,1}}
  m^n \sp{{ N \choose S+m+\mu+\nu } { N \choose S+m }}^{1/2}
  g_\mu\p{S,m+\nu} g_\nu\p{S,m}.
\end{align}

\section{The general case}

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_-^q S_z^r.
  \label{eq:H_general}
\end{align}
The Hamiltonian for one-axis twisting with a transverse-field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_z^2 + \Omega S_x
  = \chi S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{z,y}
  = \chi \p{S_z^2 - S_y^2}
  = \chi \p{S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z},
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{y,x}
  = \chi\p{S_y^2 - S_x^2}
  = -\f{\chi}{2} \p{S_+^2 + S_-^2}.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_-^q S_z^r, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n}
  = \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_-^q S_z^r,
\end{align}
where the commutator is given in Appendix \ref{sec:comm_general}, and
the decoherence operator in Appendix \ref{sec:decoherence}.  We will
be considering evolution of the initial states
$\ket{X}:S_x\ket{X}=S\ket{X}$ and $\ket{-Z}:S_z\ket{-Z}=-S\ket{-Z}$,
for which initial conditions are given by
\begin{multline}
  \bk{X|S_+^\ell S_-^m S_z^n|X}
  =  2^{-N} \sum_{k=-S+m}^{S-\ell} k^n
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}
\end{multline}
and
\begin{multline}
  \bk{-Z|S_+^\ell S_-^m S_z^n|-Z}
  = \p{-S}^n 2^{-N} \sum_{k=-S+m}^{S-\ell}
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}.
\end{multline}
Equipped with these initial conditions, we can expand
\begin{align}
  \bk{S_+^\ell S_-^m S_z^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \approx \sum_{k=0}^M \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0},
\end{align}
where $M$ is chosen such that
\begin{align}
  \f{t^M}{M!} \bk{\f{d^M}{dt^M}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \ll S^{\ell+m+n}.
\end{align}
% comment on explosion of operators
